The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-08-04 paper/referee/response-revision1.txt

MN-14-1011-MJ: Authors' response to the reviewer's comments
===========================================================

We are grateful to the referee for the insightful feedback, which has helped us
to improve the presentation of the data products.  In his/her report, the 
referee identified areas where additional information would (i) strengthen the 
role of the paper as a user's guide to the survey, and (ii) would lessen the 
burden on the reader to track down information from previous works.  
We appreciate the merits of the referee's suggestions in this respect, and have 
acted upon them by adding a total of 4 new pages (3 new sections, 3 new figures,
and 1 new table) to the manuscript.  Changes have been marked in bold.
We have also ~doubled the auxiliary documentation that is available on our web
site prompted by the referee's suggestions.  It is our intention to
expand the on-line documentation as user experience accumulates.

@@

In what follows we will start by explaining our response to each of the issues 
which the referee identified as high-level remarks (9 items), followed by the
referee's comments on specific sections (8 items).

High-level items
----------------

[Referee]
> 1. This manuscript presents a large amount of information about IPHAS and the preparation
> of and contents of the data products.  It also relies heavily on previous
> publications for general descriptions of the survey, particulars of the instrumentation,
> data processing algorithms and earlier data releases.   As such, it puts considerable burden 
> on the readers to track down information from many different sources.  It is
> understandable that the authors prefer to not duplicate information that is already
> available elsewhere, but it will make it much easier to understand and use the IPHAS data
> products if at least some of the general information  could be repeated in this
> manuscript.
> For example, it would be helpful to include concise summary of the IPHAS survey
> parameters and this data release in tabular form.  This summary would serve as
> a "IPHAS-At-A-Glance" for users, and coule include the survey area, dates, telescope,
> camera and survey bandpasses, etc.  Much of the information is presented in the paper,
> but widely spread over the manuscript.  Without at least a guide to finding the
> information, users will be challenged to find what they need.

[Authors' response]
Whilst revising the manuscript in response to the items that will follow below,
we have kept in mind the referee's recommendation to lessen the burden on the
reader.
In immediate response to this comment, we have extended Section 2 to  
(i) include an overview table of the survey parameters, which summarises 15 key 
properties of the survey, and (ii) included an extra figure to remind the reader
of the footprint of an IPHAS field pair.

@@

[Referee]
> 4. Many of the successful large surveys include a list of known caveats concerning
> released data products to help user's avoid pitfalls in their research.  I suggest that
> the authors generate a short list of the most important things about which users
> should be aware when using IPHAS data.  Users appreciate knowing how to avoid mistakes
> due to misinterpretation of known issues, and do not think less of a product because
> of them.

[Authors' response]
In response to the referee's suggestion, we have added a new Section 6.1
to alert users to the most common pitfalls.  We have also added a new
FAQ/Caveats section to our web page (www.iphas.org/dr2), which will be
updated as user experience accumulates.

[Referee]
> 5. In general, the images do not receive the same level of description in the manuscript
> as does the Catalogue.  Any guidance that can be given to users about them would be
> beneficial.  For example, at the very least include a description of the FITS image
> headers, analogous to the descriptions of the catalog columns that is given in
> the Appendix.  They should also mention that tables containing the image metadata are available at IPHAS web site.

[Authors' response]
Although the paper itself specifically targets the catalogue product, we
agree with the referee that it would be beneficial to give users
additional guidance on the use of the images.  For this reason, we
have made significant improvements to our web site by (i) providing a
query tool to retrieve IPHAS images by coordinates, 
and (ii) provide more comprehensive documentation on the metadata, the header 
keywords, the flux calibration, and the WCS astrometric solution (see www.iphas.org/images).
Moreover, Section 6 of the paper has been edited to advertise these resources.
We note that the next release, DR3, intends to focus more strongly on our image
and image mosaic products, when the survey approaches ~100% coverage.

@@

[Referee]
> 6. The fact that only 39% of the catalogue entries satisfy the criteria necessary to be
> classified as "reliable" begs the question "just what are the other 61% of objects."
> The authors should add more information about the general reliability of the
> catalogue with and without their criteria. For example, If a users selects
> "reliable" category sources, what should they expect the net reliability to be?
> Certainly not 100%.  How does reliability vary with source magnitude, SNR, or other 
> measurement parameters?  How much does the reliability vary with source environment
> (e.g. crowding, complex backgrounds)?
>
> Reliability can be gauged in a number of ways, such as comparisons with external,
> deeper catalogues (e.g. SDSS, UKIDSS GPS).

[Authors' response]
The referee questioned the nature of the objects which are marked as "reliable"
in the catalogue, and in particular questioned the nature of the large number of 
sources not flagged as "reliable".  Having considered this comment, we
have concluded that we made a poor choice of terminology by adopting the labels 
"reliable" and "veryReliable" as quality qualifiers.
These classes, which are based on stringent quality criteria set out in 
Tables 6 and 7, aim to help users identify high-quality objects, e.g. by 
demanding a 10-sigma detection in all bands.  It goes without saying that 
there are plenty of objects which do not satisfy such strict criteria but are 
nevertheless genuine astrophysical objects.  The qualifier "reliable" might 
however confer the incorrect impression that such sources are "unreliable".
As a result, we have decided to change the terminology by renaming the class 
previously known as "reliable" into "a10" (which is short for "all-band 10-sigma"), 
and renaming "veryReliable" into "a10point" (short for "all-band 10-sigma and
perfectly point-like").  We believe this change in terminology works better in
providing an operational description of the detection quality.

In addition to this change, we have adopted the referee's suggestion
to gauge the reliability of our objects against an external catalogue.  That is,
we have introduced a new Section 6.2, titled "Source reliability", in which the 
reliability of IPHAS sources as a function of the quality criteria is gauged 
against SDSS.  Moreover, we note that a separate paper is in an advanced state 
of preparation (Farnhill et al.), that investigates the issues surrounding 
the completeness and reliability of our catalogue, e.g. by injecting artificial 
sources into the images and measuring their recovery rates, whilst
also comparing source counts between SDSS and IPHAS in more detail.

@@

[Referee]
> 7. The authors have done a good job describing the photometric measurements and
> calibration accuracy.  However, there is relatively little discussion of the
> astrometric position reconstruction other than saying it is done relative to
> the 2MASS Point Source Catalogue, and there is no validation of the astrometric
> quality of the Catalogue other than to mention matching success.  Some quantification
> of the astrometric performance should be added (e.g. mean position residuals as a function
> of magnitude from comparisons with the astrometric reference stars, or external
> astrometric collections, such as UCAC4).
> Positions in the IPHAS Catalogue have been reconstructed with respect to 2MASS.
> However, the authors do not state if the proper motion of the 2MASS reference stars 
> was taken into account.  Although small, the net motion of the reference stars can
> collectively imprint on the Catalogue because the IPHAS observations were made at
> different epochs than 2MASS, in some cases as much as 12 years later.   
> Failure to do so can imprint the proper motion onto IPHAS, and also because the 
> IPHAS measurements span nearly 10 years (2003-2012), there can be secular variations.

[Authors' response]
We have edited Section 3.1 to offer the reader more details on the astrometric 
calibration.  We have also added the results of an external validation of our 
astrometry against UCAC4, which shows a mean residual of 0.09 +/- 0.07 arcsec at
the brighter end of our catalogue.  Finally, we have also added a statement
to specify that no proper motion information was used in arriving at our 
astrometry, because hundreds of 2MASS positional reference stars were typically
available in each CCD to fit the parameters of the WCS solution.

@@

[Referee]
(There does not appear to be an item number 8 in the referee report.)
> 9. The authors point out in the introduction that IPHAS will complement a number of
> new infrared surveys of the Galactic Plane.  The paper would benefit from at least
> one cross-comparison with the catalogue of one of those surveys, such as GLIMPSE
> or MIPSGAL, as an illustration of how the combination of surveys will enable
> new science.  At present, the only tie-in to the infrared surveys is the
> visual comparison between IPHAS and GLIMPSE images of one field.

[Authors' response]
In response to the referee's suggestion to include a cross-comparison between
IPHAS and an infrared survey, we have added a new section (7.3) which provides
a brief discussion on the overlap between the GLIMPSE and IPHAS catalogues.  
The discussion reminds the user that an optical survey will naturally be 
extinction-limited, by showing where sources in common between IPHAS and GLIMPSE
are positioned in a colour-colour diagram.

@@

[Referee]
> 4.a. It is stated in section 3.2 that the exposures were discarded unless they
> had 5-sigma limiting magnitudes of r>20 or i>19 and H-apha>19 mag. How were the
> limiting magnitudes for an exposure estimated?

[Authors' response]
We have added a footnote to Section 3.2 which details how the 5-sigma limiting 
magnitude is defined and estimated in our work.

[Referee]
> 4.b. Section 3.2 also specifies that an ellipticity limit of eps>0.3 was used.
> What is the justification of this limit?  For example, what is the potential impact on
> photometry and sensitivity at eps=0.25?

[Authors' response]
We have edited the third paragraph in Section 3.2 to explain that the choice
of our ellipticity limit is based on inspecting examples of high-ellipticity 
exposures, which revealed no evidence for degraded photometry up to ellipticities 
of 0.3.

@@

[Referee]
> 6.c. Section 5.3 - Table 4
> It is not clear what is the specific value is of the average magnitude given for each of
> the different band-detection combinations in Table 4?  Wouldn't something like the peak
> of the log N - log M curves be of more interest to provide some sense of the
> completeness and/or depth of the Catalogue?

[Authors' response]
Section 5.4 has been edited to make the role of Table 4 clearer.  The average
magnitudes have been removed from this table since they did not contribute to
the message which the Section attempts to convey.

[Referee]
> 6.d. Section 5.4
> While the authors provide a thorough description of the criteria used for
> Catalogue selection, no reasons were offered for why those criteria were adopted.
> For example, were there requirements for completeness and reliability that were
> being satisfies?  If not, what were the objectives?
> In the first paragraph of section 5.4, it is stated the three criteria were
> enforced to limit the size of the catalogue.  Was this the main driver for establishing
> the top-level criteria?  It seems unusual in this day and age of massive catalogues
> that this would be a driver.

[Authors' response]
The main driver for imposing the top-level criteria is to avoid the catalogue
from being dominated by the wealth of spurious detections which the pipeline
naturally picks up at S/N < 5.  We have edited the first paragraph of Section 
5.4 to clarify this point.  For DR3, we are considering following the
example set by Spitzer/GLIMPSE to offer both a "highly reliable" as
well as a  "more complete, less reliable" version of the catalogue to
the community.

@@

[Referee]
> As stated earlier in this review, some description should be offered as to the nature of
> the very large number of detections in the catalogue that are not classified as "reliable."
> It would be useful to describe them to the user, and to offer suggestions of
> how to recognize problematic detections, and also ones that may be real.

[Authors' response]
As we acknowledged during our response to "point 6", we believe that our choice
of the word "reliable" as the name of a class of high-quality objects was 
misleading. We believe we have resolved this problem by renaming the "reliable"
class as the "a10" class, which avoids conveying the incorrect impression that 
objects not in this class are spurious by default.

@@

[Referee]
> 8. Figure 11 - caption
> The reference to the left and right panels should presumably be to the top and bottom panels,
> unless the two figures will be arranged differently in the published version.

[Authors' response]
We have corrected this mistake.

---

Again, we thank the referee for the insightful feedback, which has not only
benefitted the DR2 paper, but will also be taken into account during the
preparation of DR3.

### 2014-08-04 paper/iphas-dr2-rev1.tex
Changed suggested by JED

{\bf When the previous release -- the IDR -- was created in late 2007,
just over half of the survey footprint was covered
and the data were insufficiently complete 
to support a homogeneously calibrated source catalogue.
The goal of this paper is to present the next release, 
 which supersedes the IDR by including a global re-calibration
and by taking the coverage up to 92 per cent of the survey area.}
In this work we
(i) explain the data reduction 
and quality control procedures that were applied,
(ii) describe and test the new photometric calibration, and 
(iii) detail the construction of the source catalogue
and demonstrate its use.

@@

The detailed properties of the IPHAS observing programme 
have been presented before 
by \citet{Drew2005} and \citet{Gonzalez-Solares2008}. 
To set the stage for this release,
we recap some key points in this section {\bf (see Table~1 for a summary)}.
IPHAS is an imaging survey of the Galactic Plane north of the celestial equator, 
from which photometry in Sloan $r$ and $i$ 
is extracted along with narrowband \ha. 
It is carried out using the Wide Field Camera (WFC) 
on the 2.5-metre INT in La Palma. 
It is the first digital survey to offer comprehensive optical CCD photometry
of point sources in the Galactic Plane;
the footprint spans a box 
of roughly 180 by 10 degrees, 
covering Galactic latitudes $-5\degr < b < +5\degr$ 
and longitudes $30\degr < \ell < 215\degr$.

@@

(1) \emph{Depth.} 
We discarded any exposures 
for which the $5\sigma$ limiting magnitude\footnote{\bf
We defined the $5\sigma$ limiting magnitude
as the magnitude which an imaginary point source would have
if its flux equalled five times the level of the noise in the sky background.
The sky noise was estimated using a robust MAD estimator 
for noise scaled to equivalent Gaussian standard deviation,
ie. = MAD x 1.48,
after removing large scale sky background variations.
(MAD = Median of the Absolute Deviations about the median.)}
was brighter than 20th magnitude in the $r$-band
or brighter than 19th in $i$ or \ha.
Such data were typically obtained during poor weather or full moon.
Most observations were significantly better than these limits.
Fig.~\ref{fig:depth} presents the distribution of limiting magnitudes
for all quality-approved fields;
the mean depths and standard deviations are 
$21.2\pm0.5$ ($r$), $20.0\pm0.3$ ($i$) and $20.3\pm0.3$ (\ha).
The depth achieved depended
most strongly on the presence of the moon,
which was above the horizon during 62 per cent 
of the observations.
The great range in sky brightness this
produced is behind the wide and bi-modal shape
of the $r$-band limiting magnitude distribution 
(top panel in Fig.~\ref{fig:depth}).
In contrast, the depths attained in $i$ and \ha\ 
are less sensitive to moonlight, leading to
narrower magnitude limit distributions
(middle and bottom panels in Fig.~\ref{fig:depth}).
To a lesser extent, the wide spread in the $r$-band depth
is also explained by the shorter exposure time
that was used for this band during the first months of data-taking.

(2) \emph{Ellipticity.} 
The ellipticity of a point source,
defined as $e = 1 - b / a$ 
with $b$ the semi-minor and $a$ the semi-major axis,
is a morphological measure of the elongation
of the Point Spread Function (PSF).
It is expected to be zero (circular)
in a perfect imaging system,
but it is slightly non-zero in any real telescope data 
due to optical distortions and tracking errors.
Indeed, it is worth noting
that IPHAS data have been collected
from unguided exposures that rely entirely
on the INT's tracking capability.
The mean ellipticity measured
in the data is $0.09\pm0.04$.
There have been sporadic episodes with higher ellipticities
due to mechanical glitches in the telescope tracking system.
{ \bf As a result, 3 per cent of our images show an
average ellipticity across the detectors which is worse
than $e > 0.2$.
The inspection of these examples revealed no evidence for
degraded photometry up to ellipticities less than 0.3.
In fact high ellipticities are naturally correlated with superior seeing.
We did however exclude a small number of exposures
which exceeded $e > 0.3$.}

@@

We have found the information contained
in our narrow overlap regions to be insufficient
to constrain the calibration parameters.
This is because photometry at the extreme edges of the WFC
-- where neighbouring fields overlap -- 
is prone to systematics at the level of 1--2~per cent.
The cause of these errors is thought to include 
the use of twilight sky flats in the pipeline,
which are known to be imperfect for calibrating stellar photometry 
due to stray light and vignetting \citep[e.g.][]{Manfroid1995}.
Moreover, the illumination correction in the overlap regions
is more affected by a radial geometric distortion in the WFC,
which causes the pixel scale to increase as the edges 
are approached \citep{Gonzalez-Solares2011}.
Although these systematics are reasonably small within a single field,
they can accumulate during a re-calibration process,
causing artificial zeropoint gradients across the survey
unless controlled by other external constraints.

@@

As the final step, the primary detections
selected above were compiled
into the final 99-column source catalogue
that is described in Appendix~\ref{app:columns}.
{ \bf
The original unweeded list of sources naturally included 
a significant number of spurious entries
as a result of the sensitive detection level
that is employed by the CASU pipeline.
We have decided to enforce three basic criteria
which must be met for a candidate source
to be included in the catalogue:}
\begin{enumerate}
\item the source must have been detected at S/N$>5$ in at least
one of the bands, i.e. it is required that at least one of
\emph{rErr}, \emph{iErr} or \emph{haErr} is smaller
than 0.2 mag;
\item the shape of the source must not be an obvious
cosmic ray or noise artefact, i.e. we require
either \emph{pStar} or \emph{pGalaxy} to be
greater than 20 per cent;
\item the source must not have been detected in one of the strongly
vignetted corners of the detector, 
not have had any known bad pixels in the aperture,
and not have been on the edge of one of the CCDs.
\end{enumerate}

@@

Not all the single-band detections are spurious, however.
We note that the confirmation rate for $i$-band detections
is markedly better than for $r$ and \ha,
which is likely explained by the fact that $i$ is least
affected by interstellar extinction, and so the survey can occasionally pick up
highly-reddened objects in $i$ which are otherwise lost in $r$ and \ha.
Moreover, objects which are intrinsically very red may also be picked
up in $i$ alone, while faint objects with very strong Balmer emission
may appear only in \ha.
Nevertheless, we recommend users not to rely on single-band objects
without inspecting the image data by eye,
or verifying that the object
was detected more than once (\emph{nObs}$>$1).
}

@@

Like any other photometric survey
which operates close
to the detection and resolution limits,
our catalogue inevitably contains sources
with are spurious or have been parametrised inaccurately.
In what follows, we highlight the most common
caveats which users of the catalogue might face,
followed by a discussion on how they can be avoided.
Known caveats include:
\begin{itemize}
\item \emph{Spurious objects.}
Nebulous sky backgrounds,
saturation artefacts near bright stars,
and cosmic rays
are known to be able to trigger spurious detections.
The majority of these can be removed by requiring
that a detection is made in more than one band
(\emph{nBands}$>$1),
on more than one occasion (\emph{nObs}$>$1),
or by ensuring that the object looks like a perfect
point source (\emph{pStar}$>$0.9).
\item \emph{Source blending and confusion.}
Blended objects are known to be parametrised less accurately
and to be more prone 
to source confusion
during the band-merging procedure.
We remind the reader that such objects are flagged in the catalogue
using the columns \emph{rDeblend}, \emph{iDeblend},
\emph{haDeblend}, and \emph{deblend}.
\item \emph{Low S/N detections.}
The majority of the objects in our catalogue
are faint sources observed near the detection limits,
e.g. 55 per cent of the entries in the catalogue
are fainter than $r > 20$.
The measurements of faint objects
are naturally prone to larger
random and systematic uncertainties:
for example, an inaccurately-subtracted background
will introduce a proportionally larger systematic error
for a faint object.
These objects can be removed by ensuring that an
object is detected at $S/N>10$
and has photometry which is consistent across
the different aperture sizes.
Such objects are flagged in the catalogue
using the \emph{a10} column.
\item \emph{Saturation.}
The photometry and astrometry of objects
brighter than the saturation limit of the instrument
is subject to systematic errors.
Such objects can be removed by ensuring that the columns
\emph{rSaturated}, \emph{iSaturated}, \emph{haSaturated}
or \emph{saturated} are set to ``False''.
\end{itemize}
We note that it is not possible, at this stage,
to include an exhaustive list of all the caveats,
because we cannot anticipate all forms of use of the data.
For this reason, we recommend users to read
the FAQ section on our website,
which will be updated as user experience accumulates.

@@

Most applications will require
a combination of quality criteria
to be used
to avoid the caveats outlined above.
The choice of criteria will always tension 
completeness against reliability,
i.e. the fraction of spurious sources.
To aid users we have listed two sets of
recommended quality criteria 
in Tables~\ref{tab:a10} and \ref{tab:a10point}.

First, Table~\ref{tab:a10} specifies
a set of minimum quality criteria
which should benefit most applications
which desire accurate colours
as well as completeness {\bf down to $\sim$19th magnitude}.
The listed criteria are designed to 
(i) remove low-S/N sources, 
(ii) remove saturated objects,
and (iii) remove sources for which the 2\farcs 3-diameter
aperture magnitude is inconsistent 
with its alternative 1\farcs 2-diameter measurement.
This latter criterion is a proxy
for identifying objects which are affected
by poor background subtraction
or failed deblending.
A total of 86 million sources out of 219 million 
(39 per cent) satisfy all the criteria listed in Table~\ref{tab:a10}
and are hereafter referred to as `a10'.
For convenience, the catalogue contains a boolean column
named \emph{a10} that flags these objects.

@@

For applications which require
a higher standard of {\bf accuracy} at the expense of completeness,
a set of additional quality criteria
are suggested in Table~\ref{tab:a10point}.
These criteria are designed to ensure that
(i) the object appeared as a perfect point source,
(ii) the object was not blended with a nearby neighbour,
and (iii) the object was not located near a very bright star.
59 million sources (27 per cent) satisfy
these stricter criteria 
and are hereafter referred to as `a10point'.
Again, the catalogue contains a boolean column
named \emph{a10point} which flags these objects.

@@

{\bf Towards Sh2-82,
the IPHAS and Spitzer/GLIMPSE catalogues
have 4798 entries in common,
out of 10\,739 and 11\,321 entries in total, respectively.
1356 objects of these objects were detected in all
four GLIMPSE bands.
Fig.~\ref{fig:spitzer-ccd} presents their colour-colour diagram,
in which we have indicated the 845 objects (62 per cent)
that have a counterpart in IPHAS (grey circles),
and the 511 objects that do not (red crosses).

\begin{figure}
    \includegraphics[width=\linewidth]{figures/spitzer-ccd/spitzer-ccd.pdf} 
    \caption{Spitzer/GLIMPSE colour-colour diagram towards Sh~2-82,
    showing the objects for which a counterpart exists in IPHAS
    (845 objects, grey circles) 
    and those for which no counterpart was identified 
    (511 objects, red crosses).
    A cross-matching radius of 1 arcsec
    was used to identify counterparts.
    The arrow illustrates the reddening vector corresponding
    to $A_{Ks}=3$ following the reddening law due to \citet{Flaherty2007}.
    It is apparent that objects near the top of the diagram,
    where heavily reddened objects are expected to sit,
    tend to have no counterpart.
    In contrast, many of the objects on the right hand side of the diagram,
    where YSOs are expected to sit, do have counterparts.
    %We note that three heavily-reddened objects without counterpart
    %are located above the diagram but are not shown for clarity. 
    }
    \label{fig:spitzer-ccd}
\end{figure}

We find that objects located in the upper half of the Spitzer diagram
are less likely to have an an IPHAS counterpart,
which is where highly-reddened objects are expected to sit.
This is not surprising,
because Sh2-82 is located 
towards a high-extinction sightline 
near the inner plane,
where an optical survey will naturally be extinction-limited.
Nevertheless, the extinction map due to \citet{Sale2014}
has demonstrated that IPHAS can probe stars as far away 
as 4-5\,kpc towards this region.
The situation will be even more favourable
in the outer-plane, where dust optical depths are low.

>>>>

{\bf When the previous release -- the IDR -- was created in late 2007,
just over half of the survey footprint was covered
and the data were insufficiently complete 
to support a homogeneously calibrated source catalogue.
The goal of this paper is to present the next release, 
 which supercedes the IDR by including a global re-calibration
and by taking the coverage up to 92 per cent of the survey area.}
In this work we
(i) explain the data reduction 
and quality control procedures that were applied,
(ii) describe and test the new photometric calibration, and 
(iii) detail the construction of the source catalogue
and demonstrate its use.

@@

The detailed properties of the IPHAS observing programme 
have been presented before 
by \citet{Drew2005} and \citet{Gonzalez-Solares2008}. 
To set the stage for this release,
we recap some key points in this section 
{\bf (see Table~1 for a quick-reference overview)}.
IPHAS is an imaging survey of the Galactic Plane north of the celestial equator, 
from which photometry in Sloan $r$ and $i$ 
is extracted along with narrowband \ha. 
It is carried out using the Wide Field Camera (WFC) 
on the 2.5-metre INT in La Palma. 
It is the first digital survey to offer comprehensive optical CCD photometry
of point sources in the Galactic Plane;
the footprint spans a box 
of roughly 180 by 10 degrees, 
covering Galactic latitudes $-5\degr < b < +5\degr$ 
and longitudes $30\degr < \ell < 215\degr$.

@@

(1) \emph{Depth.} 
We discarded any exposures 
for which the $5\sigma$ limiting magnitude\footnote{\bf
We defined the $5\sigma$ limiting magnitude
as the magnitude which a point source would have
if its flux equalled five times the level of the noise in the sky background.
The sky noise was estimated using a robust MAD estimator 
for noise scaled to equivalent Gaussian standard deviation,
ie. = MAD x 1.48,
after removing large scale sky background variations.
(MAD = Median of the Absolute Deviations about the median.)}
was brighter than 20th magnitude in the $r$-band
or brighter than 19th in $i$ or \ha.
Such data were typically obtained during poor weather or full moon.
Most observations were significantly better than these limits.
Fig.~\ref{fig:depth} presents the distribution of limiting magnitudes
for all quality-approved fields;
the mean depths and standard deviations are 
$21.2\pm0.5$ ($r$), $20.0\pm0.3$ ($i$) and $20.3\pm0.3$ (\ha).
The depth achieved depended
most strongly on the presence of the moon,
which was above the horizon during 62 per cent 
of the observations.
The great range in sky brightness this
produced is behind the wide and bi-modal shape
of the $r$-band limiting magnitude distribution 
(top panel in Fig.~\ref{fig:depth}).
In contrast, the depths attained in $i$ and \ha\ 
are less sensitive to moonlight, leading to
narrower magnitude limit distributions
(middle and bottom panels in Fig.~\ref{fig:depth}).
To a lesser extent, the wide spread in the $r$-band depth
is also explained by the shorter exposure time
that was used for this band during the first months of data-taking.

(2) \emph{Ellipticity.} 
The ellipticity of a point source,
defined as $e = 1 - b / a$ 
with $b$ the semi-minor and $a$ the semi-major axis,
is a morphological measure of the elongation
of the Point Spread Function (PSF).
It is expected to be zero (circular)
in a perfect imaging system,
but it is slightly non-zero in any real telescope data 
due to optical distortions and tracking errors.
Indeed, it is worth noting
that IPHAS data have been collected
from unguided exposures that rely entirely
on the INT's tracking capability.
The mean ellipticity measured
in the data is $0.09\pm0.04$.
There have been sporadic episodes with higher ellipticities
due to mechanical glitches in the telescope tracking system.
{ \bf As a result, 3 per cent of our images show an
average ellipticity across the detectors which is worse
than $e > 0.2$.
The inspection of these examples revealed no evidence
for degraded photometry up to ellipticities of 0.3.
We have excluded a small number of exposures
which exceeded $e = 0.3$.}

@@

We have found the information contained
in our narrow overlap regions to be insufficient
to constrain the calibration parameters well enough.
This is because photometry at the extreme edges of the WFC
-- where neighbouring field pairs overlap -- 
is prone to systematics at the level of 1--2~per cent.
The cause of these errors is thought to include 
the use of twilight sky flats in the pipeline,
which are known to be imperfect for calibrating stellar photometry 
due to stray light and vignetting \citep[e.g.][]{Manfroid1995}.
Moreover, the illumination correction in the overlap regions
is more affected by a radial geometric distortion in the WFC,
which causes the pixel scale to increase as the edges 
are approached \citep{Gonzalez-Solares2011}.
Although these systematics are reasonably small within a single field,
they can accumulate during a re-calibration process,
causing artificial zeropoint gradients across the survey
unless controlled by other external constraints.

@@

As the final step, the primary detections
selected above were compiled
into the final 99-column source catalogue
that is described in Appendix~\ref{app:columns}.
{ \bf
The original unweeded list of sources naturally included 
a significant number of spurious entries
as a result of the sensitive detection level
that is employed by the CASU pipeline.
We have decided to enforce three basic criteria
which must be met for a candidate source
to be included in the catalogue:}
\begin{enumerate}
\item the source must have been detected at S/N$>5$ in at least
one of the bands, i.e. it is required that at least one of
\emph{rErr}, \emph{iErr} or \emph{haErr} is smaller
than 0.2 mag;
\item the shape of the source must not be an obvious
cosmic ray or noise artefact, i.e. we require
either \emph{pStar} or \emph{pGalaxy} to be
greater than 20 per cent;
\item the source must not have been detected
in one of the strongly vignetted corners of the instrument, 
not have had any known bad pixels in the aperture,
and not have been on the edge of one of the CCDs.
\end{enumerate}

@@

Not all the single-band detections are spurious, however.
We note that the confirmation rate for $i$-band detections
is markedly better than for $r$ and \ha,
which is likely to be explained by the fact that $i$ is least
affected by interstellar extinction, and so the survey can occasionally pick up
highly-reddened objects in $i$ which are otherwise lost in $r$ and \ha.
Moreover, objects which are intrinsically very red may also be picked
up in $i$ alone, while faint objects with very strong Balmer emission
may appear only in \ha.
Nevertheless, we recommend users not to rely on single-band objects
without inspecting the image data by eye,
or verifying that the object
was detected more than once (\emph{nObs}$>$1).
}

@@

Like any other photometric survey
in which a majority of detected objects are close
to the detection and resolution limits,
our catalogue inevitably contains sources
that are spurious or have been parametrised inaccurately.
In what follows, we highlight the most common
caveats which users of the catalogue might face,
followed by a discussion on how they can be avoided.
These caveats include:
\begin{itemize}
\item \emph{Spurious objects.}
Nebulous sky backgrounds,
saturation artefacts near bright stars,
and cosmic rays
are known to be able to trigger spurious detections.
The majority of these can be removed by requiring
that a detection is made in more than one band
(\emph{nBands}$>$1),
on more than one occasion (\emph{nObs}$>$1),
or by ensuring that the object looks like a perfect
point source (\emph{pStar}$>$0.9).
\item \emph{Source blending and confusion.}
Blended objects are known to be parametrised less accurately
and to be more prone 
to source confusion
during the band-merging procedure.
We remind the reader that such objects are flagged in the catalogue
using the columns \emph{rDeblend}, \emph{iDeblend},
\emph{haDeblend}, and \emph{deblend}.
\item \emph{Low S/N detections.}
The majority of the objects in our catalogue
are faint sources observed near the detection limits,
e.g. 55 per cent of the entries in the catalogue
are fainter than $r > 20$.
The measurements of faint objects
are naturally prone to larger
random and systematic uncertainties:
for example, an inaccurately-subtracted background
will introduce a proportionally larger systematic error
for a faint object.
These objects can be removed by ensuring that an
object is detected at $S/N>10$
and has photometry which is consistent across
the different aperture sizes.
Objects detected at $S/N>10$ in all bands
are flagged in the catalogue using the \emph{a10} column.
\item \emph{Saturation.}
The photometry and astrometry of objects
brighter than the saturation limit of the instrument
is subject to systematic errors.
Such objects can be removed by ensuring that the columns
\emph{rSaturated}, \emph{iSaturated}, \emph{haSaturated}
or \emph{saturated} are set to ``False''.
\end{itemize}
We note that it is not possible, at this stage,
to include an exhaustive list of all the caveats,
because we cannot anticipate all forms of use of the data.
For this reason, we recommend users to read
the FAQ section on our website,
which will be updated as user experience accumulates.

@@

Many applications will require
a combination of quality criteria
to be used
to avoid the issues identified above.
The choice of criteria will always tension 
completeness against reliability,
i.e. the fraction of spurious sources.
To aid users we have listed two sets of
recommended quality criteria 
in Tables~\ref{tab:a10} and \ref{tab:a10point}.

First, Table~\ref{tab:a10} specifies
a set of minimum quality criteria
which should benefit most applications
which desire accurate colours
as well as completeness down to $\sim$19th magnitude.
The listed criteria are designed
to ensure that each band offers photometry
at a high S/N which is consistent
across different aperture sizes.
This latter criterion is a proxy
for identifying objects which are affected
by poor background subtraction
or failed deblending.
A total of 86 million sources out of 219 million 
(39 per cent) satisfy all the criteria listed in Table~\ref{tab:a10}
and are hereafter referred to as `a10'
(short for ``all-band $10\sigma$'').
For convenience, the catalogue contains a boolean column
named \emph{a10} that flags these objects.

@@

For applications which require
a higher standard of accuracy at the expense of completeness,
a set of additional quality criteria
are suggested in Table~\ref{tab:a10point}.
These criteria are designed to ensure that
(i) the object appeared as a perfect point source in all bands,
(ii) the object was not blended with a nearby neighbour,
and (iii) the object was not located near a very bright star.
59 million sources (27 per cent) satisfy
these stricter criteria 
and are hereafter referred to as `a10point'.
Again, the catalogue contains a boolean column
named \emph{a10point} which flags these objects.

@@

{\bf Towards Sh2-82,
the IPHAS and Spitzer/GLIMPSE catalogues
have 4798 entries in common,
out of 10\,739 and 11\,321 entries in total, respectively.
Fig.~\ref{fig:spitzer-ccd} presents the Spitzer colour-colour diagram
of the region, containing the 1356 objects that were detected 
in all four GLIMPSE bands.
We have indicated the 845 objects (62 per cent)
that have a counterpart in IPHAS (grey circles)
and the 511 objects that do not (red crosses).

\begin{figure}
    \includegraphics[width=\linewidth]{figures/spitzer-ccd/spitzer-ccd.pdf} 
    \caption{Spitzer/GLIMPSE colour-colour diagram towards Sh~2-82,
    showing the objects for which a counterpart exists in IPHAS
    (845 objects, grey circles) 
    and those for which no counterpart was identified 
    (511 objects, red crosses).
    A cross-matching radius of 1 arcsec
    was used to identify counterparts.
    The arrow illustrates the reddening vector corresponding
    to $A_{Ks}=3$ following the reddening law due to \citet{Flaherty2007}.
    It is apparent that objects near the top of the diagram,
    where heavily reddened objects are expected to sit,
    tend to have no counterpart.
    In contrast, many of the objects on the right hand side of the diagram,
    where YSOs are expected to sit, do have counterparts.
    }
    \label{fig:spitzer-ccd}
\end{figure}

We find that objects located in the upper half of the Spitzer diagram
are less likely to have an IPHAS counterpart,
which is where highly-reddened objects are expected to sit.
This is not surprising,
because Sh2-82 is located 
towards a high-extinction sightline 
near the inner plane,
where an optical survey will naturally be extinction-limited.
Nevertheless, the extinction map due to \citet{Sale2014}
has demonstrated that IPHAS can probe stars as far away 
as 4-5\,kpc towards this region.
The situation will be even more favourable
in the outer-plane, where dust optical depths are low.
